Modeling Ionic Polymer-Metal Composites\\ with Space-Time Adaptive Multimesh \emph{hp}-FEM

### 2011-02-25 response.tex

\item \emph{The authors have not given the initial values of $C$}

The initial value was added below Eq.~(1.1).

\item \emph{The authors have not given the size of time-step and the relation
of time step size and h; p or CFL stability condition.}

All computations performed in this paper were done using the absolutely stable 
second-order Crank-Nicolson method. This method does not require any 
limitation of time step such as the CFL condition. Of course, the 
accuracy is still an issue and the size of the time step was chosen 
based on multiple experiments, so that the size of the temporal error 
was approximately the same as the error in space. The ultimate answer 
to this issue is adaptive time stepping which is the subject of a forthcoming 
paper. We have given the size of the time step, and 
we explained in more detail how the choice of the fixed time step
for the Crank-Nicolson method was made in the very beginning
of Section 5.

### 2011-02-25 intro.tex
Revision1 done

Ionic Polymer-Metal Composites (IPMC) have been studied during the past two 
decades for their potential to serve as noiseless mechanoelectrical and electromechanical 
transducers \cite{basu1997membrane,shahinpoor2001smartmat,
nasser2002applied,newbury2003intelligent, wallmersperger2007appliedphysics,
pugal2008appliedphysics,pugal2010polymer}.
The advantages of IPMC over other electroactive polymer actuators
are low voltage bending, high strains ($>1\%$), and an ability to work
in wet environments. A typical IPMC consists of a thin sheet of polymer
(often Nafion or Teflon) which is sandwiched between noble
metal electrodes such as platinum or gold. When fabricated, the polymer 
membrane is saturated with certain solvent and ions such as water and $H^+$.
When a voltage is applied to the electrodes, the counter ions start
migrating due to the imposed electric field. By dragging along the solvent,
the osmotic pressure difference near the electrodes
results in bending of the material (see Fig.~\ref{fig:conceptual}).
\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[scale=0.7]{IPMC_bending}
  \caption{\label{fig:conceptual}Conceptual model of the actuation
 	of IPMC. Initial counter ion distribution (a) and
	the distribution and resulting bending after applying a voltage (b).}
  \end{centering}
%\vspace{-1cm}
\end{figure}
%\newpage
%The derived and implemented model helps to predict the actuation of the
%material. Furthermore, it is expected that the \emph{hp}-FEM implementation
%results in a smaller problem size, thus likely allowing faster calculation
%in both 2D and 3D domains.

>>>>

Ionic Polymer-Metal Composites (IPMC) have been studied during the past two 
decades for their potential to serve as noiseless mechanoelectrical and electromechanical 
transducers \cites{basu1997membrane,shahinpoor2001smartmat,
nasser2002applied,newbury2003intelligent, wallmersperger2007appliedphysics,
pugal2008appliedphysics,pugal2010polymer}.
The advantages of IPMC over other electroactive polymer actuators
are low voltage bending, high strains ($>1\%$), and an ability to work
in wet environments. A typical IPMC consists of a thin sheet of polymer
(often Nafion or Teflon) which is sandwiched between noble
metal electrodes such as platinum or gold. When fabricated, the polymer 
membrane is saturated with certain solvent and ions such as water and $H^+$.
When a voltage is applied to the electrodes, the counter ions start
migrating due to the imposed electric field. By dragging along the solvent,
the osmotic pressure difference near the electrodes
results in bending of the material (see Fig.~\ref{fig:conceptual}).
\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[scale=0.7]{IPMC_bending}
  \caption{\label{fig:conceptual}Conceptual model of the actuation
 	of IPMC. Initial counter ion distribution (a) and
	the distribution and resulting bending after applying a voltage (b).}
  \end{centering}
%\vspace{-1cm}
\end{figure}
%\newpage
%The derived and implemented model helps to predict the actuation of the
%material. Furthermore, it is expected that the \emph{hp}-FEM implementation
%results in a smaller problem size, thus likely allowing faster calculation
%in both 2D and 3D domains.

### 2011-02-25 model.tex
Revision1 done

\section{Model}\label{sec:model}

@@

conditions for $\phi$, we obtain

@@

The $hp$-FEM is a modern version of the Finite Element Method (FEM) that is 
capable of exponential convergence (the approximation error drops exponentially 
as new degrees of freedom are added during adaptivity) while standard FEM can 
only attain algebraic (polynomial) convergence rates which are much slower
[REFERENCE SOLIN-BOOK-2003].

>>>>

\newpage
\section{Model}\label{sec:model}

@@

conditions for $\varphi$, we obtain

@@

The $hp$-FEM is a modern version of the Finite Element Method (FEM) that is 
capable of exponential convergence (the approximation error drops exponentially 
as new degrees of freedom are added during adaptivity) while standard FEM can 
only attain algebraic (polynomial) convergence rates which are much slower~\cite{Hermes-book}.

### 2011-02-25 results.tex
Changes by Pavel.
Revision1 done

The solutions to the PNP problem exhibit a specific behavior that was 
described above. In order to find the best adaptive method to deal with 
this type of problems, we performed numerous computations using all 
adaptivity modes in both the single-mesh and multi-mesh regimes.
In the numerical experiments we paid attention to the 
relative error, cumulative CPU time, and problem size 
in terms of number of degrees of freedom (DOF) in each 
time step. The scaled variables $c$
and $\varphi$ and the unscaled time $t$ are used to present the solutions.
The simulations were performed with physical time step of $0.05$~s and
the final time of $3.0$~s was chosen as it is close to the time
scaling constant~$\tau$.

@@

To optimize the calculation time of HP\_ANISO, an adaptive time step control was employed.
The classical PID controller was used~\cite{valli2002control,dubcova2010space}.
Since $c$ and $\varphi$ change differently in time as was demonstrated in Figs.~\ref{fig:cphi-1}
and~\ref{fig:cphi-2}, the relative changes between the solutions at different
time steps were monitored:
\begin{eqnarray}
  e_c^n & = & \frac{\lVert c^n-c^{n-1}\rVert}{\lVert c^n \rVert},\\
  e_\varphi^n & = & \frac{\lVert \varphi^n-\varphi^{n-1}\rVert}{\lVert \varphi^n \rVert}.
\end{eqnarray}
The relative changes to control the time step was calculated as follows:
\begin{equation}
  e^n=\max\left\{ e_c^n,\ e_\varphi^n \right\}.
\end{equation}
If $e^n < \delta$ where $\delta>0$ is a defined tolerance, then the time
step for the next iteration is increased smoothly to
\begin{equation}
  \delta\tau^{n+1}=\left( \frac{e^{n-1}}{e^n} \right)^{k_P}\ \left( \frac{\delta}{e^n} \right)^{k_l}\
  \left[ \frac{\left( e^{n-1} \right)^2}{e^n e^{n-2}} \right]^{k_D} \delta\tau^n,
\end{equation}
where parameters are from~\cite{valli2002control}:
\begin{equation}
  k_p=0.075,\quad k_l=0.175,\quad k_D=0.01.
\end{equation}
The tolerance $\delta$ was set to $\delta=0.25$ in the current optimization
example. At this point, the implementation does not support
reducing the time step when $e^n\geq \delta$. The calculated $e_c^n$ and $e_\varphi^n$ are shown in Fig.~\ref{fig:cphirelerr}.
The HP\_ANISO problem size and computing time with and without time step control
are shown in Figs.~\ref{fig:timeadapt_dof} and~\ref{fig:timeadapt_cpu}. The reader
can notice that the computing time was reduced more than two times when the time
step control was employed.

>>>>

The solutions to the PNP problem exhibit a specific behavior that was 
described above. In order to find the best adaptive method to deal with 
this type of problems, we performed numerous computations using all 
adaptivity modes in both the single-mesh and multi-mesh regimes.
In the numerical experiments we paid attention to the 
relative error, cumulative CPU time, and problem size 
in terms of number of degrees of freedom (DOF) in each 
time step. The scaled variables $c$
and $\varphi$ and the unscaled time $t$ are used to present the solutions.
The simulations were performed with physical time step of $0.05$~s and
the final time of $3.0$~s was chosen as it is close to the time
scaling constant~$\tau$.  The time step was chosen after many
numerical experiments in such a way that the error
in time was approximately the same as the error in
space. The implementation of advanced adaptive
implicit higher-order Runge-Kutta methods is in progress.

@@

To optimize the calculation time of HP\_ANISO, an adaptive time step control was employed.
The classical PID controller was used~\cite{valli2002control,dubcova2010space}.
Since $c$ and $\varphi$ change differently in time as was demonstrated in Figs.~\ref{fig:cphi-1}
and~\ref{fig:cphi-2}, the relative changes between the solutions at different
time steps were monitored:
\begin{eqnarray}
  e_c^n & = & \frac{\lVert c^n-c^{n-1}\rVert}{\lVert c^n \rVert},\\
  e_\varphi^n & = & \frac{\lVert \varphi^n-\varphi^{n-1}\rVert}{\lVert \varphi^n \rVert}.
\end{eqnarray}
The relative changes to control the time step was calculated as follows:
\begin{equation}
  e^n=\max\left\{ e_c^n,\ e_\varphi^n \right\}.
\end{equation}
If $e^n < \delta$ where $\delta>0$ is a defined tolerance, then the time
step for the next iteration is increased smoothly to
\begin{equation}
  \delta\tau^{n+1}=\left( \frac{e^{n-1}}{e^n} \right)^{k_P}\ \left( \frac{\delta}{e^n} \right)^{k_l}\
  \left[ \frac{\left( e^{n-1} \right)^2}{e^n e^{n-2}} \right]^{k_D} \delta\tau^n,
\end{equation}
where parameters are from~\cite{valli2002control}:
\begin{equation}
  k_p=0.075,\quad k_l=0.175,\quad k_D=0.01.
\end{equation}
The tolerance $\delta$ was set to $\delta=0.25$ in the current optimization
example. At this point, the implementation does not support
adaptive time stepping if $e^n\geq \delta$. However, the implementation 
of advanced adaptive higher-order time-stepping methods is in progress.
The calculated $e_c^n$ and $e_\varphi^n$ are shown in Fig.~\ref{fig:cphirelerr}.
The HP\_ANISO problem size and computing time with and without time step control
are shown in Figs.~\ref{fig:timeadapt_dof} and~\ref{fig:timeadapt_cpu}. The reader
can notice that the computing time was reduced more than two times when the time
step control was employed.
